TBD

### 2019-09-13 manuscript/manuscript.Rmd
edits before sending to Jordi; also incorporated Jordi's edits into Impact Summary.

\vspace*{0.1cm} 
\begin{center} \LARGE Consumer extinctions constrain phenotypic evolution in the resulting food web \end{center}

@@

*Manuscript elements*: table 1, table 2, figure 1, figure 2, figure 3, figure 4. All figures should be printed in color.

@@

*Keywords*: adaptive landscape, ecological networks, eco-evolutionary dynamics, natural selection, host-parasitoid; complexity.

@@

*Number of words*: `r wordcountaddin::word_count("manuscript.Rmd")`

@@

The loss of biodiversity<!--Global change--> is altering the structure of ecological networks; however, we are currently in a poor position to predict how these altered communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that simulated the extinction of a guild of larval parasitoids that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that the number of traits under selection increased with the extinction of larval parasitoids. In contrast, the adaptive landscape was more neutral in the original food web because different parasitoid guilds impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that the loss of trophic interactions due to consumer extinctions can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes to phenotypes to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory], and gives insight to the direct and indirect effects of selective constraints [@Arnold1992]. Ecological interactions often play a key role in shaping selective constraints, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change. Although there is clear evidence that species interactions can shape the adaptive landscape, we still have a poor understanding of how the adaptive landscape is shaped by a community of interacting species [@McPeek2017; @terHorst2018]. Resolution on how ecological communities shape phenotypic evolution is urgently needed though, given the rapid losses of biodiversity we are observing in the Anthropocene [@Scheffers2016].

Ecological networks, such as food webs describing who-eats-whom in ecological communities, provide an explicit representation of species interactions in an ecological community [@Bascompte2014; @McCann2012]. Here, we integrate ecological networks and adaptive landscapes to understand how ecological communities constrain evolutionary change. The direct and indirect effects of selective constraints can be inferred by quantifying the slope and curvature of the adaptive landscape [@Arnold1992]. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Selective constraints on evolution increase with the number of traits under directional selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape is governed by nonlinear selection gradients and acts to indirectly constrain evolution through its effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. Stabilizing selection acts to erode genetic variance in traits, which can impose a constraint on the ability of those traits to respond to future selection [@Hansen2008].<!-- In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008].--> Correlational selection may alter the genetic covariance between traits, which may facilitate or hinder future adaptation depending on pattern of selection on those traits [@Hansen2008].<!--Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008].--> If we want to predict how ecological communities shape phenotypic evolution, we must understand how ecological networks shape the adaptive landscape.

The loss of biodiversity is altering the structure of ecological networks<!--communities-->, which may influence evolutionary constraints in a number of ways. For example, in a <!--multitrophic community or -->food web, if different consumers impose directional selection on different traits of a shared resource, then a more diverse consumer community may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more diverse communities. To examine these different possibilities, we conducted a field experiment that simulated the extinction of a consumer guild associated with an abundant insect herbivore (*Iteomyia salicisverruca*, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web structure by either excluding the guild of larval parasitoids (extinction food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (original food web, fig. \ref{fig:Conceptual}). The relative abundance of these larval parasitoids is much lower than the egg parasitoid [@Barbour2016], and thus represents a plausible extinction scenario for this system. <!--While this manipulation simulates an extreme co-extinction event, the likelihood of such co-extinction events is increasing in the Anthropocene. -->Note that larval parasitoids also impose indirect effects on gall midge fitness through intraguild predation on the egg parasitoid. We then applied modern statistical methods to quantify how changes in food-web structure altered the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how the loss of biodiversity may constrain the evolution of interacting populations.

```{r Conceptual, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Conceptual}Experimental manipulation of food-web structure associated with a leaf-galling midge (*Iteomyia salicisverruca*) feeding on the willow *Salix hookeriana*. Black arrows denote the flow of energy in this network of trophic interactions. In the original food web, we allowed the full suite of egg and larval parasitoids to impose selection. To simulate consumer extinction, we used mesh bags to exclude the guild of larval parasitoids, only allowing the egg parasitoid (*Platygaster* sp.) to impose selection. Note that larval parasitoids also impose indirect effects on gall midge fitness through intraguild predation on the egg parasitoid. Larval parasitoids include the following species (from left to right): *Mesopolobus* sp. (Family: Pteromalidae); *Tetrastichus* sp. (Family: Eulophidae); and *Torymus* sp. (Family: Torymidae)."}
knitr::include_graphics("../analyses/complex_simple_foodwebs_v3.jpeg")
```

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. For the original food web (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that all galls had similar phenology when we collected galls later in the season. To simulate consumer extinctions, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989]. It was not possible to reciprocally manipulate parasitoid attack (i.e. exclude egg parasitoid, but allow larval parasitoids), because it was not possible to identify midge oviposition sites prior to gall formation. In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. We did not include other sources of mortality, such as early larval death, as they could influence the expression of the gall phenotype and confound estimates of selection. <!--Since we were interested in selection imposed by parasitoids, we excluded other sources of mortality (`r round((mean(full_df$early.larval.death) + mean(full_df$unknown_mortality)), 2)*100`% mortality from other sources). Another reason why we excluded other sources of mortality, such as early larval death, is that they can influence the expression of the gall phenotype and confound estimates of selection.--> For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

Inferring adaptive landscapes assumes that trait distributions are multivariate normal [@Lande1983]. 
To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. We then scaled all phenotypic traits (mean=0 and SD=1) prior to our analyses (detailed below) to ensure that our estimates of selection were comparable across traits and with other studies.

Our analysis consisted of four parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and standardized phenotypic traits ($i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate standardized selection gradients. Third, we used our estimates of selection gradients to characterize the slope and curvature of the adaptive landscape, which we used to measure evolutionary constraints. Finally, we used a simulation to examine how the curvature of the adaptive landscape could alter the evolvability of populations in the next generation.

**Selection surfaces**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_i$), quadratic ($\alpha_{ii}$), and linear interactions ($\alpha_{ij}$) between each trait. We also allowed these selection surfaces ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear selection surfaces, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think that modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e., setting them to zero) when calculating selection surfaces. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as selection surfaces in each food-web treatment. <!--To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.-->

**Selection gradients**: Selection gradients cannot be estimated directly from GLMMs of selection surfaces because the response is in terms of absolute fitness and the coefficients are on a nonlinear scale. For example, the coefficients in the above model measure the change in the logarithm of the odds of surviving (i.e., $\text{ln}\{W(z)/[1-W(z)]\}$) associated with 1 SD change in a trait with all other traits held fixed at their mean. Therefore, we used the method developed by @Janzen1998 to translate selection surfaces from the above model into the scale of relative fitness in order to calculate directional ($\beta_i$), quadratic ($\gamma_{ii}$), and correlational ($\gamma_{ij}$) selection gradients. Briefly, this method calculates the average gradient of selection surfaces by multiplying the average of $W(z)[1-W(z)]$ by each regression coefficient (e.g. $\alpha_i$, $\alpha_{ii}$, or $\alpha_{ij}$). We then divided this average gradient by the mean fitness ($\bar W$) to put it on the scale of relative fitness, and thus interpretable as a selection gradient. We estimated selection gradients separately for each food-web treatment. Note that we doubled all quadratic terms prior to calculating selection gradients to put them on the same scale as estimates of directional and correlational selection [@Stinchcombe2008].

**Evolutionary constraints**: We measured evolutionary constraints by inspecting the slope and curvature of the adaptive landscape. The number of selective constraints is determined by the slope of the adaptive landscape, which in our study corresponds to:

$$\textbf{Slope} = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
where each $\beta_i$ corresponds to the directional selection gradient acting on each trait. By comparing the number of directional selection gradients that show clear evidence of contributing to the slope (i.e., 95%  CI does not overlap zero) in our food-web treatments, we can infer the effect of food-web structure on selective constraints.

The indirect effects of selection on the G-matrix ($\Delta\text{G}=\text{G}(\gamma - \beta \beta^\text{T})\text{G}$) is governed by the curvature of the adaptive landscape ($\text{C}=\gamma - \beta \beta^\text{T}$), which in our study corresponds to:

@@

$$\textbf{Curvature} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}$$
where each $\text{C}_{ii}$ (diagonal) represents the effect of selection on the additive genetic variance in a trait, and each $\text{C}_{ij}$ (off-diagonal) represents the effect of selection on the additive genetic covariance between a particular trait combination. In other words, the sign of diagonal terms of the curvature matrix dictate whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) indicate the effects of selection on additive genetic covariance between two traits. We used bootstrapped values of each selection gradient to estimate the curvature of each component of the matrix and its associated 95% CI. We also used this information to determine whether the curvature of each component differed between our food-web treatments.

Knowledge of the curvature matrix alone gives an incomplete picture of its indirect effect on genetic constraints. This is because genetic constraints are ultimately determined by the structure of the G-matrix, and thus may depend on its structure before selection. Although we do not know the underlying G-matrix for the traits we measured in this experiment, we can still gain insight to how our food-web treatment would alter genetic constraints. Specifically, we first calculated how our best estimate of the curvature matrix (mean values) in each treatment changed the structure of `r n_Gmats` random G-matrices. If we assume that recombination/mutation would have the same effect on gall midge populations in the original and extinction food webs, then we can estimate how the curvature matrix in the previous generation alters the G-matrix for the next generation.

The G-matrix itself is a complex structure, but recent theory has shown how its structure affects the evolvability of a population [@Hansen2008]. <!--We used evolvability as our measure of genetic constraint, as it has a solid foundation in theory of quantitative genetics and adaptive landscapes [@Hansen2008]. -->Evolvability measures the ability for a trait to evolve toward a given direction of selection, ignoring responses in every other direction [@Hansen2008]. In the absence of knowledge about the direction of selection that a population will actually experience in the next generation, we can assess its evolutionary potential by computing its average evolvability over random selection gradients (here, we used `r n_random_gradients`). By computing the average evolvability for each of `r n_Gmats` random G-matrices, we can then assess how changes in the curvature matrix alter the evolutionary potential of a population. We report the distribution of these effect sizes, rather than conduct a statistical test, because the number of replicates in a simulation can be arbitrarily high, thus making any effect size statistically significant [@White2014].

@@

Once parasitized, the gall phenotype becomes the phenotype of the egg parasitoid. This phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify selection surfaces and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Consumer extinctions increase selective constraints

>>>>

\vspace*{0.1cm} 
\begin{center} \LARGE Loss of consumers constrains phenotypic evolution in the resulting food web \end{center}

@@

*Short Running Title*: Consumer loss constrains phenotypic evolution

@@

*Keywords*: adaptive landscape, ecological networks, eco-evolutionary dynamics, natural selection, host-parasitoid; complexity; biodiversity.

@@

*Abstract Word Count*: 251

*Total Word Count*: `r wordcountaddin::word_count("manuscript.Rmd") - 587` <!-- subtract estimate of Abstract, Impact, Acknowledgements, Author Contributions, and Data Accessibility -->

@@

The loss of biodiversity<!--Global change--> is altering the structure of ecological networks; however, we are currently in a poor position to predict how these altered communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the network context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that removed<!--simulated the extinction of--> a guild of larval parasitoids that imposed direct and indirect selection pressures on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits to estimate directional, quadratic, and correlational selection gradients in each treatment. We used these selection gradients to characterize the slope and curvature of the adaptive landscape to understand the direct and indirect effects of consumer loss on evolutionary constraints. We found that the number of traits under directional selection increased with the removal of larval parasitoids, indicating greater selective constraints on the trajectory of evolutionary change. Similarly, we found that the removal of larval parasitoids altered the curvature of the adaptive landscape in such a way that tended to decrease the evolvability of the traits we measured in the next generation.<!--In contrast, the adaptive landscape was more neutral in the original food web because different parasitoid guilds impose different selection pressures, minimizing relative fitness differences among phenotypes.--> Our results suggest that the loss of trophic interactions <!--due to consumer extinctions -->can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

# Impact Summary

<!-- Authors should provide an approximately 300 word summary, explaining why the article makes an important contribution to Evolutionary Biology. It should be accessible to a wide audience, e.g. journalists, non-expert public interested in evolution, school science teachers. -->

The loss of biodiversity is rewiring the web of life; however, we are currently in a poor position to predict how this will affect the ability of populations to evolve and adapt to future environments. To gain insight to these effects, we conducted a field experiment that either maintained a natural community of predators or only allowed one of the predators to impose selection on a common prey. We found that the loss of predators increased the number of prey traits under selection. This suggests that simpler communities may constrain the trajectory of evolution toward a specific combination of traits. Moreover, we found that that the loss of predators could make it more difficult for prey to adapt to future environments. Taken together, our results suggest that the loss of biodiversity may rewire interactions in such a way that constrains the adaptive potential of remaining populations.

@@

The adaptive landscape provides a powerful framework for understanding how natural selection has shaped the evolution of biodiversity ---from genes to phenotypes to species [@Wright1931; @Simpson1944; @Arnold2001]. More than a metaphor, the adaptive landscape links quantitative genetic and phenotypic variation to evolution by natural selection [@Lande1979; @Arnold1984applications; @Arnold1984theory], giving insight to the direct and indirect effects of selective constraints [@Arnold1992]. Ecological interactions often play a key role in shaping selective constraints, as evidenced by the role of antagonistic [@Schluter2000; @Abrams2000] and mutualistic [@Bronstein2006] interactions in driving evolutionary change. Although there is clear evidence that species interactions can shape the adaptive landscape, we still have a poor understanding of how the adaptive landscape is shaped by a community of interacting species [@McPeek2017; @terHorst2018]. Resolution on how change in ecological communities shape phenotypic evolution is urgently needed though, given the rapid losses of biodiversity we are observing in the Anthropocene [@Scheffers2016].

Ecological networks, such as food webs describing who-eats-whom, provide an explicit representation of the direct and indirect effects that emerge in a community of interacting species [@Bascompte2014; @McCann2012]. Here, we integrate ecological networks and adaptive landscapes to understand how ecological communities constrain evolutionary change. The direct and indirect effects of natural selection can be inferred by quantifying the slope and curvature of the adaptive landscape [@Arnold1992]. For example, the slope is determined by directional selection gradients acting on each phenotypic trait and influences the trajectory of evolutionary change [@Lande1979;@Arnold1992]. Selective constraints on evolution increase with the number of traits under directional selection, as this diminishes the number of optimal solutions [@Arnold2003]. The curvature of the adaptive landscape is governed by nonlinear selection gradients and acts to indirectly constrain evolution through its effect on genetic constraints [@Arnold1992; @Hansen2008]. Genetic constraints are largely governed by a population's **G**-matrix ---the additive genetic variances and covariances between traits [@Hansen2008]. Stabilizing selection acts to erode genetic variance in traits, which can impose a constraint on the ability of those traits to respond to future selection [@Hansen2008].<!-- In general, genetic constraints will increase with the number of traits under directional or stabilizing selection, as this will decrease the additive genetic variance in those traits [@Hansen2008].--> Correlational selection may alter the genetic covariance between traits, which may facilitate or hinder future adaptation depending on pattern of selection on those traits [@Hansen2008].<!--Genetic constraints may also increase with the number of trait combinations under correlational selection, as this type of selection decreases the evolutionary independence of traits [@Hansen2008].--> If we want to predict how ecological communities shape phenotypic evolution, we must understand how ecological networks shape the adaptive landscape.

The loss of biodiversity is altering the structure of ecological networks<!--communities-->, which may influence evolutionary constraints in a number of ways. For example, in a <!--multitrophic community or -->food web, if different consumers impose directional selection on different traits of a shared resource, then a more diverse consumer community may constrain evolution by increasing the number of traits under selection. Alternatively, if consumers impose selection on different values of a trait, then their selective effects would cancel each other out in more diverse communities. To examine these different possibilities, we conducted a field experiment that removed a consumer guild that directly and indirectly impacts fitness of an abundant insect herbivore (*Iteomyia salicisverruca*, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped galls when they feed on the developing leaves of willow trees [*Salix* sp., @Russo2006]. These galls protect larva from attack by generalist predators (e.g. ants, spiders), but they suffer high mortality from egg and larval parasitoids [@Barbour2016]. We manipulated food-web structure by either removing larval parasitoids (removal food web) or allowing both egg and larval parasitoids to impose selection on gall midge traits (original food web, fig. \ref{fig:Conceptual}). The relative abundance of these larval parasitoids is much lower than the egg parasitoid [@Barbour2016], and thus represents a plausible extinction scenario for this system. <!--While this manipulation simulates an extreme co-extinction event, the likelihood of such co-extinction events is increasing in the Anthropocene. -->Larval parasitoids also impose indirect effects on gall midge fitness through intraguild predation on the egg parasitoid (fig. \ref{fig:Conceptual}). We applied modern statistical methods to quantify how changes in food-web structure altered the slope and curvature of the gall midge's adaptive landscape. Taken together, our study gives insight to how the loss of biodiversity may constrain the evolution of interacting populations.

```{r Conceptual, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:Conceptual}Experimental manipulation of food-web structure associated with a leaf-galling midge (*Iteomyia salicisverruca*) feeding on the willow *Salix hookeriana*. Black arrows denote the flow of energy in this network of trophic interactions. In the original food web, we allowed the full suite of egg and larval parasitoids to impose selection. In the removal food web, we used mesh bags to exclude the guild of larval parasitoids, only allowing the egg parasitoid (*Platygaster* sp.) to impose selection. Note that larval parasitoids also impose indirect effects on gall midge fitness through intraguild predation on the egg parasitoid. Larval parasitoids include the following species (from left to right): *Mesopolobus* sp. (Family: Pteromalidae); *Tetrastichus* sp. (Family: Eulophidae); and *Torymus* sp. (Family: Torymidae)."}
knitr::include_graphics("../analyses/complex_simple_foodwebs_v3.jpeg")
```

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. For the original food web (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that all galls had similar phenology when we collected galls later in the season. For the removal food web, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989]. It was not possible to reciprocally manipulate parasitoid attack (i.e. exclude egg parasitoid, but allow larval parasitoids), because it was not possible to identify midge oviposition sites prior to gall formation. In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. We did not include other sources of mortality, such as early larval death, as they could influence the expression of the gall phenotype and confound estimates of selection. <!--Since we were interested in selection imposed by parasitoids, we excluded other sources of mortality (`r round((mean(full_df$early.larval.death) + mean(full_df$unknown_mortality)), 2)*100`% mortality from other sources). Another reason why we excluded other sources of mortality, such as early larval death, is that they can influence the expression of the gall phenotype and confound estimates of selection.--> For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

Inferring adaptive landscapes assumes that trait distributions are multivariate normal [@Lande1983]. 
To approximate this assumption, we log-transformed clutch size and square-root transformed oviposition preference. We then scaled all phenotypic traits (mean=0 and SD=1) across treatments prior to our analyses (detailed below) to ensure that our estimates of selection were comparable across traits and with other studies.

Our analysis consisted of four parts. First, we used generalized linear mixed models (GLMM) to quantify selection surfaces \textemdash linear and nonlinear relationships between absolute fitness ($W$) and standardized phenotypic traits ($i$) of individuals \textemdash in each food-web treatment. Second, we translated selection surfaces into the scale of relative fitness ($w$) in order to calculate standardized selection gradients. Third, we used our estimates of directional selection gradients to characterize the slope of the adaptive landscape, which we used to measure selective constraints. Finally, we estimated the curvature of the adaptive landscape and used a simulation to explore its effects on the adaptive potential of the gall midge population in the next generation.

**Selection surfaces**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_i$), quadratic ($\alpha_{ii}$), and linear interactions ($\alpha_{ij}$) between each trait. We also allowed these selection surfaces ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear selection surfaces, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think that modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e. setting them to zero) when calculating selection surfaces. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as selection surfaces in each food-web treatment. To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.

**Selection gradients**: Selection gradients cannot be estimated directly from GLMMs of selection surfaces because the response is in terms of absolute fitness and the coefficients are on a nonlinear scale. For example, the coefficients in the above model measure the change in the logarithm of the odds of surviving (i.e. $\text{ln}\{W(z)/[1-W(z)]\}$) associated with 1 SD change in a trait with all other traits held fixed at their mean. Therefore, we used the method developed by @Janzen1998 to translate selection surfaces from the above model into the scale of relative fitness in order to calculate directional ($\beta_i$), quadratic ($\gamma_{ii}$), and correlational ($\gamma_{ij}$) selection gradients. Briefly, this method calculates the average gradient of selection surfaces by multiplying the average of $W(z)[1-W(z)]$ by each regression coefficient (e.g. $\alpha_i$, $\alpha_{ii}$, or $\alpha_{ij}$). We then divided this average gradient by the mean fitness ($\bar W$) to put it on the scale of relative fitness, and thus interpretable as a selection gradient. We estimated selection gradients separately for each food-web treatment. We also determined whether selection gradients differed between food-web treatments by calculating the difference in bootstrapped replicates between treatments. Note that we doubled all quadratic terms prior to calculating selection gradients to put them on the same scale as estimates of directional and correlational selection [@Stinchcombe2008].

**Selective constraints**: <!--We measured evolutionary constraints by inspecting the slope and curvature of the adaptive landscape. -->The number of selective constraints is determined by the slope of the adaptive landscape, which in our study corresponds to:

$$\textbf{Slope} = \begin{pmatrix} \beta_{\text{Diam}} \\ \beta_{\text{Clutch}} \\ \beta_{\text{Pref}} \end{pmatrix} $$
where each $\beta_i$ corresponds to the directional selection gradient acting on each trait. By comparing the number of directional selection gradients that show clear evidence of contributing to the slope (i.e. 95%  CI does not overlap zero) in our food-web treatments, we can infer the effect of food-web structure on selective constraints. This gives insight as to whether there is a specific combination of phenotypes favored by natural selection (i.e. more traits under directional selection), or whether multiple trait combinations have comparable fitness (i.e. less traits under directional selection).

**Genetic constraints**: The indirect effects of selection on the G-matrix ($\Delta\text{G}=\text{G}(\gamma - \beta \beta^\text{T})\text{G}$) is governed by the curvature of the adaptive landscape ($\text{C}=\gamma - \beta \beta^\text{T}$), which in our study corresponds to:

@@

$$\textbf{Curvature} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}$$
where each $\text{C}_{ii}$ (diagonal) represents the effect of selection on the additive genetic variance in a trait, and each $\text{C}_{ij}$ (off-diagonal) represents the effect of selection on the additive genetic covariance between a particular trait combination. <!--In other words, the sign of diagonal terms of the curvature matrix dictate whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) indicate the effects of selection on additive genetic covariance between two traits. -->We used bootstrapped values of each selection gradient to estimate the curvature of each component of the matrix and its associated 95% CI. We also used this information to determine whether the curvature of each component differed between our food-web treatments.

Knowledge of the curvature matrix alone gives an incomplete picture of its indirect effect on genetic constraints. This is because genetic constraints are ultimately determined by the structure of the G-matrix, and therefore also depend on its structure before selection. Although we do not know the underlying G-matrix for the traits we measured in this experiment, we can still gain insight to how our food-web treatment would alter genetic constraints more generally. Specifically, we calculated how our best estimate of the curvature matrix (mean values) in each treatment changed the structure of `r n_Gmats` random G-matrices for the next generation. We restricted the range of additive genetic variance (*V~G~*) in these G-matrices to between 0 and 0.5 to reflect typical ranges in narrow-sense heritability values (*h*^2^; note that *h*^2^ = *V~G~* when the phenotypic variance is scaled to 1). Note that this analysis assumes that the effects of recombination and mutation on the G-matrix are much smaller than the effects of selection, which appears to be a reasonable assumption over short time scales [@Arnold2008].

The G-matrix itself is a complex structure, but has a clear theoretical link to the evolvability of phenotypic traits [@Hansen2008]. <!--We used evolvability as our measure of genetic constraint, as it has a solid foundation in theory of quantitative genetics and adaptive landscapes [@Hansen2008]. -->Evolvability measures the ability for a trait to evolve toward a given direction of selection [@Hansen2008]. In the absence of knowledge about the direction of selection that a population will actually experience in the next generation, we can compute its average evolvability over random selection gradients [@Hansen2008; @Melo2015]. By computing the average evolvability (here, we used `r n_random_gradients` random $\beta$s) for each of `r n_Gmats` random G-matrices, we can then assess how changes in the curvature matrix alter the evolutionary potential of the associated traits. We report the distribution of these effect sizes, rather than conduct a statistical test, because the number of replicates in a simulation can be arbitrarily high, thus making any effect size statistically significant [@White2014].

@@

Once parasitized, the gall phenotype becomes the phenotype of the egg parasitoid. This phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify selection surfaces and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the loss of consumers may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Consumer loss increases selective constraints

\indent We found that the removal of larval parasitoids increased the number of gall midge traits under directional selection (3 of 3) relative to the original food web (1 of 3)(table \ref{Table:Gradients}). For example, we observed directional selection for smaller clutch sizes when we excluded larval parasitoids, but there was no evidence of selection acting on this trait in the original food web (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of selection appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids. Specifically, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes (larval parasitoids $\beta_{\text{Clutch}}$=
`r round(filter(tidy_egglarval_grads, type=="Original", term=="sc.log.Clutch")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Original", term=="sc.log.Clutch")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Original", term=="sc.log.Clutch")$conf.high,2)`]). This conflicting selection is likely due to trait differences between guilds, as larger clutches may be easier targets for egg parasitoids to find, but the more complex gall structure may be more difficult for larval parasitoids to oviposit through.
